sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-22 doc/paper.Rmd
Edit abstract

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], Other authors\n| $^1$Pacific Biological Station Fisheries and Oceans Canada, \n| Nanaimo, BC, Canada\n|
  $^2$Other addresses\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Analyzing geostatistical data---spatially referenced observations---is increasingly common in ecology for applications such as species distribution models (SDMs). These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2: We introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models, with an interface familiar to those who have used lme4 or glmmTMB. Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB).
    | 3: We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the package to related software. We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a user-friendly interface familiar to many R users, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
    | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, Template Model Builder
---

@@

Here, we introduce the R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv, for a specific class of spatial and spatiotemporal models [@wood2017a].
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

>>>>

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], Other authors\n| $^1$Pacific Biological Station Fisheries and Oceans Canada, \n| Nanaimo, BC, Canada\n|
  $^2$Other addresses\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Analyzing geostatistical data---spatially referenced observations---is increasingly common in ecology for applications such as species distribution models (SDMs). These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2: We introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models. Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB). We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the package to related software.
    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
    | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, Template Model Builder
---

@@

Here, we introduce the R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv [@wood2017a], for a specific class of spatial and spatiotemporal models.
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of the R package and the underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.
